{\bf Practical Concurrency Testing} \\ \normalsize \vspace{1em} \begin{tabular}{c} {\em or: How I Learned to Stop Worrying and Love the Exponential Explosion} \end{tabular}

### 2018-12-24 thesis/overview.tex
assorted minor proofreads
consistency of function ()s in overview

\begin{figure}[t]
	\begin{tabular}{cc}
		\begin{tabular}{p{0.45\textwidth}p{0.5\textwidth}}
			{\small
			\begin{tabular}{l}
				\texttt{\ctype{int} x = \const{0};} \\
				\texttt{\ctype{void} \call{count}() \{} \\
				\texttt{~~~~\call{for} (\ctype{int} i = \const{0}; i < \const{1000}; i++)} \\
				\texttt{~~~~~~~~x++;} \\
				\texttt{\}} \\
				\texttt{\ctype{void} \call{main}() \{} \\
				\texttt{~~~~tid1 = \call{thr\_create}(count);} \\
				\texttt{~~~~tid2 = \call{thr\_create}(count);} \\
				\texttt{~~~~\call{thr\_join}(tid1);} \\
				\texttt{~~~~\call{thr\_join}(tid2);} \\
				\texttt{~~~~\call{printf}(\const{"\%d\textbackslash{}n"}, x);} \\
				\texttt{\}} \\
			\end{tabular}
			}
			&
			{\small \begin{tabular}{ll}
				{\bf Thread 1} & {\bf Thread 2} \\
				\hline
				\texttt{load tmp <- x;} & \\
				& \texttt{load tmp <- x;} \\
				& \texttt{add tmp <- 1;} \\
				& \texttt{store x <- tmp;} \\
				\texttt{add tmp <- 1;} & \\
				\texttt{store x <- tmp;} & \\
			\end{tabular}
			}
			\\
			\\
			(a) Source listing for a multithreaded program which might count to 2000.
			&
			% revisionminor can't go here i guess
			(b) Example interleaving of (a)'s compiled assembly,
			in which two concurrent iterations of the loop yield one net increment of {\tt x}.
		\end{tabular}
	\end{tabular}
	\caption[Example concurrent program.]{Example concurrent program in which simultaneous accesses to shared state may interleave to produce unexpected results.}
	\label{fig:concurrency-bug}
\end{figure}

@@

\section{Stateless Model Checking}
\label{sec:overview-stateless} % ugh this label name

@@

The thread lifecycle/management routines are as follows:
\begin{itemize}
	\item {\tt thr\_init(stack\_size)} initializes the thread library, setting a \revisionminor{custom, fixed} stack size to be allocated to new threads.
	\item {\tt thr\_create(child\_func, child\_arg)} spawns a new thread to run the specified function with the specified argument. There is a semantic gap between this function and the {\tt thread\_fork} system call (which takes no parameters, makes no changes to the user's address space, and cannot meaningfully be invoked from C code) which students must bridge.
		Returns an integer thread ID of the newly created thread.
	\item {\tt thr\_exit(status)} \revisionminor{completes} execution of the calling thread, recording an exit status value.
		\revisionminor{This function does not return to its callsite.}
		The main challenge of this function is to allow another thread to free the memory used for the exiting thread's stack,
		without risking any corruption as long as the exiting thread continues to run.
	\item {\tt thr\_join(tid, statusp)} blocks the calling thread until the thread with the specified thread ID exits, then returns, collecting its exit status.
\end{itemize}
Other than {\tt thr\_init} (which is necessarily single-threaded), several concurrency errors between any two (or all three) of these functions are very common in student submissions.

@@

The Pintos kernel architecture \cite{pintos} is used at several universities, including the University of Chicago, Berkeley, and Stanford.
Its basecode implements a rudimentary kernel, consisting of a context switcher, round-robin scheduler, locking primitives, and program loader,
upon which students add more features in several projects.
Most relevant to this thesis, the basecode provides the following functions/libraries, among others:
\begin{itemize}
	\item Semaphores (the basic concurrency primitive, implemented using direct scheduler calls): {\tt sema\_up}, {\tt sema\_down}, {\tt sema\_try\_down};
	\item Locks (which wrap a semaphore initialized to 1), {\tt lock\_acquire}, {\tt lock\_\allowbreak{}release}, {\tt lock\_try\_acquire};
	\item Condition variables (also implemented using scheduler calls): {\tt cond\_wait}, {\tt cond\_\allowbreak{}signal}, {\tt cond\_broadcast}, with the same semantics as Pebbles P2 condvars;
	\item Basic round-robin scheduling facilities: {\tt thread\_block} (a kernel-level analogue to Pebbles's {\tt deschedule}), {\tt thread\_yield}
	\item Kernel thread lifecycle management, {\tt thread\_create} and {\tt thread\_exit}, including stack space memory management;
	\item Interrupt and fault handlers;
	\item A page allocator, {\tt palloc\_get\_page}, {\tt palloc\_get\_multiple}, {\tt palloc\_free\_page}, and {\tt palloc\_free\_multiple}
\end{itemize}
Both Pebbles and Pintos basecodes offer a standard C library including {\tt malloc}, string-formatting, printing, et cetera.

>>>>

\begin{figure}[t]
	\begin{tabular}{cc}
		\begin{tabular}{p{0.44\textwidth}p{0.495\textwidth}}
			{
			\begin{tabular}{l}
				\texttt{\ctype{int} x = \const{0};} \\
				\texttt{\ctype{void} \call{count}() \{} \\
				\texttt{~~~~\call{for} (\ctype{int} i = \const{0}; i < \const{1000}; i++)} \\
				\texttt{~~~~~~~~x++;} \\
				\texttt{\}} \\
				\texttt{\ctype{void} \call{main}() \{} \\
				\texttt{~~~~tid1 = \call{thr\_create}(count);} \\
				\texttt{~~~~tid2 = \call{thr\_create}(count);} \\
				\texttt{~~~~\call{thr\_join}(tid1);} \\
				\texttt{~~~~\call{thr\_join}(tid2);} \\
				\texttt{~~~~\call{printf}(\const{"\%d\textbackslash{}n"}, x);} \\
				\texttt{\}} \\
			\end{tabular}
			}
			&
			{\begin{tabular}{ll}
				{\bf Thread 1} & {\bf Thread 2} \\
				\hline
				\texttt{load tmp <- x;} & \\
				& \texttt{load tmp <- x;} \\
				& \texttt{add tmp <- 1;} \\
				& \texttt{store x <- tmp;} \\
				\texttt{add tmp <- 1;} & \\
				\texttt{store x <- tmp;} & \\
			\end{tabular}
			}
			\\
			\\
			(a) Source listing for a multithreaded program which might count to 2000.
			&
			% revisionminor can't go here i guess
			(b) Example interleaving of (a)'s compiled assembly,
			in which two concurrent iterations of the loop yield one net increment of {\tt x}.
		\end{tabular}
	\end{tabular}
	\caption[Example concurrent program.]{Example concurrent program in which simultaneous accesses to shared state may interleave to produce unexpected results.}
	\label{fig:concurrency-bug}
\end{figure}

@@

\section{Stateless model checking}
\label{sec:overview-stateless} % ugh this label name

@@

The thread lifecycle/management routines are as follows:
\begin{itemize}
	\item {\tt thr\_init(stack\_size)} initializes the thread library, setting a \revisionminor{custom, fixed} stack size to be allocated to new threads.
	\item {\tt thr\_create(child\_func, child\_arg)} spawns a new thread to run the specified function with the specified argument. There is a semantic gap between this function and the {\tt thread\_fork} system call (which takes no parameters, makes no changes to the user's address space, and cannot meaningfully be invoked from C code) which students must bridge.
		Returns an integer thread ID of the newly created thread.
	\item {\tt thr\_exit(status)} \revisionminor{completes} execution of the calling thread, recording an exit status value.
		\revisionminor{This function does not return to its callsite.}
		The main challenge of this function is to allow another thread to free the memory used for the exiting thread's stack,
		without risking any corruption as long as the exiting thread continues to run.
	\item {\tt thr\_join(tid, statusp)} blocks the calling thread until the thread with the specified thread ID exits, then returns, collecting its exit status.
\end{itemize}
Other than {\tt thr\_init()} (which is necessarily single-threaded),
concurrency errors between any two (or all three) of these functions are very common in student submissions.

@@

The Pintos kernel architecture \cite{pintos} is used at several universities, including the University of Chicago, Berkeley, and Stanford.
Its basecode implements a rudimentary kernel, consisting of a context switcher, round-robin scheduler, locking primitives, and program loader,
upon which students add more features in several projects.
Most relevant to this thesis, the basecode provides the following functions/libraries, among others:
\begin{itemize}
	\item Semaphores (the basic concurrency primitive, implemented using direct scheduler calls): {\tt sema\_up()}, {\tt sema\_down()}, {\tt sema\_try\_down()};
	\item Locks (which wrap a 1-initialized semaphore), {\tt lock\_acquire()}, {\tt lock\_\allowbreak{}release()}, {\tt lock\_try\_acquire()};
	\item Condition variables (also implemented using scheduler calls): {\tt cond\_wait()}, {\tt cond\_\allowbreak{}signal()}, {\tt cond\_broadcast()}, with the same semantics as Pebbles P2 condvars;
	\item Basic round-robin scheduling facilities: {\tt thread\_block()} (a kernel-level analogue to Pebbles's {\tt deschedule}), {\tt thread\_yield()}
	\item Kernel thread lifecycle management, {\tt thread\_create()} and {\tt thread\_exit()}, including stack space memory management;
	\item Interrupt and fault handlers;
	\item A page allocator, {\tt palloc\_get\_page()}, {\tt palloc\_get\_multiple()}, {\tt palloc\_free\_page()}, and {\tt palloc\_free\_multiple()}.
\end{itemize}
Both Pebbles and Pintos basecodes offer a standard C library including {\tt malloc}, string-formatting, printing, et cetera.
